---
title: "Exercise - Gradient Descent"
author: "Qi"
date: "2025-04-17"
output: 
  pdf_document:
    number_sections: true
    toc: true
    toc_depth: 2
    latex_engine: xelatex
---
\newpage
# Introduction

In this exercise, we explore the concept of **gradient descent**, a fundamental optimization algorithm used to minimize functions. We will apply it to a simple one-dimensional function and approximate the derivative using a secant slope (finite differences) instead of requiring the user to provide the analytical derivative.

# What is Gradient Descent?

Gradient descent is an iterative optimization algorithm used to find the local minimum of a differentiable function. The idea is to start from an initial guess and move in the direction opposite to the gradient (i.e., the direction of steepest descent), scaled by a learning rate.

## The General Update Rule:

Given a function \( f(x) \), the update rule for gradient descent is:

\[ x_{t+1} = x_t - \eta \nabla f(x_t) \]

Where:

  - \( x_t \) is the current estimate of the minimum
  - \( \eta \) is the learning rate (step size)
  - \( \nabla f(x_t) \) is the derivative of \( f \) at \( x_t \)

In this exercise, we approximate the derivative numerically.

# Function to Minimize

We will apply gradient descent to the following function:

\[ f(x) = x + \frac{1}{x}, \quad x > 0 \]

This function is smooth and has a unique minimum on the domain \( (0, \infty) \).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Step-by-Step Implementation in R

## Step 1: Define the Function

```{r}
f <- function(x) {
  y <- ifelse(x > 0, x + 1 / x, Inf)  # handle vectorized input
  return(y)
}
```

## Step 2: Implement the Gradient Descent Algorithm with Numerical Derivative

```{r}
gradient_descent <- function(f, x0, eta = 0.01, tol = 1e-6, max_iter = 1000, h = 1e-6) {
  x <- x0
  iter <- 0
  path <- numeric(max_iter)

  repeat {
    # numerical derivative via finite differences
    grad <- (f(x + h) - f(x - h)) / (2 * h)
    x_new <- x - eta * grad
    if (x_new <= 0) x_new <- x / 2  # keep x in (0, ∞)

    path[iter + 1] <- x_new
    if (abs(x_new - x) < tol || iter >= max_iter) {
      break
    }
    x <- x_new
    iter <- iter + 1
  }

  list(
    x_min = x_new,
    f_min = f(x_new),
    iterations = iter,
    path = path[1:iter]
  )
}
```

## Step 3: Run the Algorithm

```{r}
result <- gradient_descent(f, x0 = 2, eta = 0.05)

cat("Minimum point:", result$x_min, "\n")
cat("Minimum value:", result$f_min, "\n")
cat("Iterations:", result$iterations, "\n")
```

## Step 4: Visualize the Optimization Path

```{r}
curve(f, from = 0.5, to = 2.5, col = "blue", lwd = 2, ylab = "f(x)", main = "Gradient Descent Path")
points(result$path, f(result$path), col = "red", pch = 20)
points(result$path[length(result$path)], f(result$path)[length(result$path)], col = "green", pch = 20, lwd = 2)
text(result$path[1:20], f(result$path[1:20]), labels = seq_along(result$path[1:20]), pos = c(1,3), cex = 0.5, col = "black")

text(result$path[length(result$path)], f(result$path[length(result$path)]), labels = length(result$path), pos = c(3), cex = 0.5, col = "black")


```

# Discussion

- The gradient descent algorithm converged to a unique minimum near \( x = `r round(result$x_min, 4)` \).
- Because the function is convex on \( (0, \infty) \), the algorithm reliably converges regardless of the initial point (as long as \( x > 0 \)).
- We used numerical differentiation instead of an analytical gradient to make the implementation more general.

# Conclusion

This exercise illustrates how gradient descent works with numerical gradients on a convex function. Understanding the step-by-step behavior helps build intuition for tuning parameters and diagnosing optimization issues in more complex models.
